---
title: Analysing data jobs in industry
output:
    github_document:
        html_preview: true
---
<br>

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

Data analysts and data scientists are in demand in many industries around the world. Since getting my first analyst job a couple of years ago I've wondered what really sets these roles (and the people who do them) apart. Here I'll use the **StackOverflow 2019 Developer Survey** to explore the characteristics of data analysts and data scientists, including **salary**, **contribution to open source**, **coding languages used**, and **coding as a hobby**. 

<br>

This project was inspired by Julia Silge's amazing analysis of [gender and salary in the tech industry](https://juliasilge.com/blog/salary-gender/) which I highly recommend! 

The pre-processing steps are executed in *prepare_data.r*. To summarise, the data are prepared as follows:

* salary above \$30,000 and below \$2,000,000
    
* full-time workers only
    
* academics and upper management removed

* filtered to self-identified data analysts or data scientists

#### What do data analysts and data scientists earn?



```{r, fig.width=10}
source("prepare_data.r")
source("plot_data.r")

salary_grid_1
```

