---
title: "**Analysing careers in data with the StackOverflow Developer Survey 2019**"
output:
    github_document:
        html_preview: true
---

<br>

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

This project was inspired by Julia Silge's amazing analysis of [gender and salary in
the tech industry](https://juliasilge.com/blog/salary-gender/). There's a lot of cool
analysis that can be done with the survey data, but I'll be focussing my analysis on
"data" jobs - namely data analysts and data scientists. 

### Pre-processing

The pre-processing steps are executed in *prepare_data.r*. To summarise, respondents are filtered as:
* salary above $30,000 and below $2,000,000
* full-time only
* self-identified data analyst or data scientist

### What's the difference between data analysts and data scientists?

```{r}
source("prepare_data.r")

library(tidyverse)

data_jobs_UKIR %>%
    ggplot(aes(ConvertedComp, colour = DevType, fill = DevType)) +
    geom_density(alpha = 0.4, size = 1) +
    scale_x_log10(labels = scales::dollar_format()) +
    theme_minimal() +
    labs(
        x = "Salary (USD)",
        y = "Density",
        fill = "Developer type",
        colour = "Developer type"
    )

```

