---
title: "**Analysing careers in data with the StackOverflow Developer Survey 2019**"
output:
    github_document:
        html_preview: true
---
<br>

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

Data analysts and data scientists are in demand across industries. Usually the difference between these jobs is that data scientists use more machine learning and programming than analysts, but this boundary can sometimes become blurred. Here I'll use the StackOverflow 2019 Developer Survey to explore
the characteristics of data analysts and data scientists, including **salary**, **contribution to open source**, and **coding as a hobby**. 

<br>

This project was inspired by Julia Silge's amazing analysis of [gender and salary in the tech industry](https://juliasilge.com/blog/salary-gender/) which I highly recommend! 

The pre-processing steps are executed in *prepare_data.r*. To summarise, respondents are filtered as:

* salary above q\$30,000 and below \$2,000,000
    
* full-time only
    
* self-identified data analyst or data scientist

#### What do data analysts and data scientists earn?

```{r, fig.width=10}
source("prepare_data.r")
source("plot_data.r")

salary_grid
```

