---
title: Analysing data jobs in industry
output:
    github_document:
        html_preview: true
---
<br>

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
source("prepare_data.r")
```

In this project I'll be taking a look at what data scientists and data analysts do differently in their day-to-day work and how this relates to their salary. The data comes from **StackOverflow's 2019 Developer Survey** which a great resource for finding out about people who develop professionally or as a hobby. 
<br>

This project was inspired by Julia Silge's analysis of [gender and salary in the tech industry](https://juliasilge.com/blog/salary-gender/) which I highly recommend! 

#### Preparing the data

To fit a more accurate model of salary the data were filtered to:

* full-time workers 

* salaries between between \$30,000 and \$2,000,000 (salaries below this are unlikely to be full-time, salaries above are most likely threshold values; see the [survey methodology](https://insights.stackoverflow.com/survey/2019#methodology))

* individual contributors working in industry only (i.e. removing upper management and academics)

Respondents were labelled as data analysts if they identified as a "Data or business analyst" or a data scientist if identifying as "Data scientist or machine learning specialist". `r nrow(df_both_jobs)` respondents identified with both labels. These responses were removed to focus the analysis on the differences between the two groups. This left `r nrow(df_one_job)` responses. The salaries used here are the converted US Dollary salaries provided by StackOverflow. 

#### What do data analysts and data scientists earn?

If we take a look at salaries overall, the range of salaries is similar between the groups with a large positive skew. 

```{r, fig.width=10}
source("plot_data.r")
salary_grid_1
```

The summary statistics show that the median salary for data scientists is slightly higher than for data analysts. 

```{r}
df_one_job %>%
    group_by(DevType) %>%
    summarise(`Median salary` = median(ConvertedComp),
              `Min salary`  = min(ConvertedComp),
              `Max salary` = max(ConvertedComp)) %>%
    knitr::kable()
```
